%!TEX root = EJIR_title_page.tex

\documentclass[times]{sagej}

\begin{document}

\title{Firm-level bargaining and within-firm wage inequality: Evidence across Europe}

\date{23 February 2021}

\author{Valeria Cirillo\affilnum{1} and Federico Tamagni\affilnum{2} and Matteo Sostero \affilnum{2,3}}

\affiliation{%
\affilnum{1} Department of Political Sciences, University of Bari “Aldo Moro”, Bari, Italy\\
\affilnum{2} Institute of Economics, Sant'Anna School of Advanced Studies, Pisa, Italy\\
\affilnum{3} Joint Research Center, European Commission, Seville. The scientific output expressed here does not imply a policy position of the European Commission.
Neither the European Commission nor any person acting on behalf of the Commission is responsible for the use which might be made of this publication.
}

\corrauth{Federico Tamagni, c/o Institute of Economics, Scuola Superiore Sant'Anna, Piazza Martiri della libertà 33, 56127, Pisa, Italy}
\email{f.tamagni@sssup.it}

\runninghead{Cirillo, Tamagni, Sostero}

\begin{abstract}
The collective bargaining process in European economies has gradually shifted from centralized bargaining to firm-level agreements over the last decades. This transition allows firms to change wages based on local market conditions, potentially affecting the wages of the highly-paid and the low-paid employees differently, with uncertain overall effects on within-firm wage inequality. We test whether firm-level bargaining is associated with greater overall within-firm wage inequality using linked employer-employee earnings data in 2006 and 2010 for six European economies, representing different collective bargaining traditions in Europe: Belgium, Spain, Germany, France, the Czech Republic and the UK. Controlling for employee and firm characteristics, we find that within-firm inequality -- measured as the gap between the 90th and 10th percentile of wage residuals in each firm -- was narrower under firm-level bargaining than in centralized bargaining in the UK, but it widened over time in France and Spain.
\end{abstract}

\keywords{within-firm wage inequality, firm-level bargaining, coordinated bargaining, matched employer-employee data}

\maketitle

\begin{dci}
The authors declare no competing interests.
This work was primarily conducted at the Sant'Anna School of Advanced Studies.
The scientific output expressed here does not imply a policy position of the European Commission.
Neither the European Commission nor any person acting on behalf of the Commission is responsible for the use which might be made of this publication.
\end{dci}

\end{document}
